DynaProg for Scala

### 2013-11-05 docs/report/conclusion.tex
Nits.

We have demonstrated that our implementation, DynaProg, is able to deal with simple (Smith-Waterman) and complex (Zuker and RNAfold) dynamic programming problems on sequences that are expressed by a grammar and an algebra. With this formalism, their expression is simplified and exploiting the dynamic programming results becomes possible thanks to systematic backtrack encoding that allows traces to be exchanged between algebrae. This implementation is publicly available\footnote{\url{https://github.com/manojo/lamp-dp-mt/}} and can immediately be used in Scala projects.

>>>>

We have demonstrated that our implementation, DynaProg, is able to deal with simple (Smith-Waterman) and complex (Zuker and RNAfold) dynamic programming problems on sequences that are expressed by a grammar and an algebra. With this formalism, their expression is simplified and exploiting the dynamic programming results becomes possible thanks to systematic backtrack encoding that allows traces to be exchanged between algebras. This implementation is publicly available\footnote{\url{https://github.com/manojo/lamp-dp-mt/}} and can immediately be used in Scala projects.

### 2013-11-05 docs/report/design.tex
Nits.

Note that this optimization needs to use informations from both the grammar (the rule to split) and the algebra ($f$), which restricting its application to either single algebra or algebra that could share the same optimization, because modifying the grammar will change the associated backtrack information (thereby breaking its compatibility for usage with other algebrae).

@@

LMS only supports a subset of Scala. Embedding both algebrae and grammars into the LMS framework would reduce the expressivity of algebrae. Taking the example of matrix multiplication, the user would like to not only solve the dynamic problem on GPU, but also leverage the graphic card, the disk or the network to compute the matrices products. Since these resources are outside for the LMS world, the user needs to write an ad-hoc DSL for every particular function. In order to avoid this constraint, we need to decouple the grammar and algebra generation.

For the generation of algebra functions, since CUDA types are restricted, they are representable in LMS. Hence we can leverage LMS for CUDA-compatible algebrae whereas regular Scala can be used for complex algebrae being executed on CPU exclusively.

>>>>

Note that this optimization needs to use informations from both the grammar (the rule to split) and the algebra ($f$), which restricting its application to either single algebra or algebra that could share the same optimization, because modifying the grammar will change the associated backtrack information (thereby breaking its compatibility for usage with other algebras).

@@

LMS only supports a subset of Scala. Embedding both algebras and grammars into the LMS framework would reduce the expressivity of algebras. Taking the example of matrix multiplication, the user would like to not only solve the dynamic problem on GPU, but also leverage the graphic card, the disk or the network to compute the matrices products. Since these resources are outside for the LMS world, the user needs to write an ad-hoc DSL for every particular function. In order to avoid this constraint, we need to decouple the grammar and algebra generation.

For the generation of algebra functions, since CUDA types are restricted, they are representable in LMS. Hence we can leverage LMS for CUDA-compatible algebras whereas regular Scala can be used for complex algebras being executed on CPU exclusively.

### 2013-11-05 docs/report/implementation.tex
Nits.

% ----------------------------------------------
\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
First define a signature that can fit both algebrae, then specify for each algebra the related functions. In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general).
\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignSignature extends Signature {
  type Alphabet = Char
  def start(x:Unit):Answer
  def gap1(g:(Int,Int),a:Answer):Answer
  def gap2(a:Answer,g:(Int,Int)):Answer
  def pair(c1:Alphabet,a:Answer,c2:Alphabet):Answer
}

@@

Finally, we describe the associated grammar and the programs that mixes the algebrae and the grammar. Note that we need one instance of each pair of grammar and algebra. Once we have done that, we can request scores and backtracks associated with an evaluation algebra (Smith-Waterman or Needleman-Wunsch) and reuse the obtained backtrack to construct the matching aligned sequences:
\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
  val axiom:Tabulate = tabulate("M",(
    empty                     ^^ start
  | seq1() -~ axiom           ^^ gap1
  |           axiom ~- seq2() ^^ gap2
  | el1    -~ axiom ~- el2    ^^ pair
  ) aggregate h)
}

@@

Although the whole program can be defined in a single trait, it is preferable to cleanly separate the signature from the grammar and the algebra, this good practice would help adding new algebrae easily. The signature needs to inherit either from {\tt Signature} or {\tt RNASignature}, would the RNA folding energies be needed. The grammar can be either single track or two-tracks by inheriting respectively from  {\tt ADPParsers} and {\tt TTParsers}. Note that RNA folding only works for single track grammars and library setup is enabled by the usage of the trait {\tt RNASignature} (this could be changed by disabling the flag {\tt energies}).

>>>>

% ----------------------------------------------
\subsubsection{Smith-Waterman and Needleman-Wunsch} \label{ex_swat}
First define a signature that can fit both algebras, then specify for each algebra the related functions. In this example, both algebra operate on the same output domain and share the same optimization function (although this is not true in general).
\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignSignature extends Signature {
  type Alphabet = Char
  def start(x:Unit):Answer
  def gap1(g:(Int,Int),a:Answer):Answer
  def gap2(a:Answer,g:(Int,Int)):Answer
  def pair(c1:Alphabet,a:Answer,c2:Alphabet):Answer
}

@@

Finally, we describe the associated grammar and the programs that mixes the algebras and the grammar. Note that we need one instance of each pair of grammar and algebra. Once we have done that, we can request scores and backtracks associated with an evaluation algebra (Smith-Waterman or Needleman-Wunsch) and reuse the obtained backtrack to construct the matching aligned sequences:
\begin{lstlisting}[language=Scala,captionpos=none]
trait SeqAlignGrammar extends TTParsers with SeqAlignSignature {
  val axiom:Tabulate = tabulate("M",(
    empty                     ^^ start
  | seq1() -~ axiom           ^^ gap1
  |           axiom ~- seq2() ^^ gap2
  | el1    -~ axiom ~- el2    ^^ pair
  ) aggregate h)
}

@@

Although the whole program can be defined in a single trait, it is preferable to cleanly separate the signature from the grammar and the algebra, this good practice would help adding new algebras easily. The signature needs to inherit either from {\tt Signature} or {\tt RNASignature}, would the RNA folding energies be needed. The grammar can be either single track or two-tracks by inheriting respectively from  {\tt ADPParsers} and {\tt TTParsers}. Note that RNA folding only works for single track grammars and library setup is enabled by the usage of the trait {\tt RNASignature} (this could be changed by disabling the flag {\tt energies}).

### 2013-11-05 docs/report/introduction.tex
Nits.

A \textbf{signature} serves as interface between the grammar, the scoring algebra and the aggregation function, which allows the user to combine one grammar with different algebrae or vice versa. Because recurrence relations are expressed by a parsing grammar, ADP makes the candidate structure explicit and hides tabulation indices, thereby preventing potential errors. Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and an algebra, it becomes possible to systematically convert dynamic programming descriptions into efficient recurrences for many-core platforms such as GPUs \cite{adp_gpu}.

@@

Practically, an ADP program is made of 3 components: a \textbf{signature} that define a set of function signatures, one or more \textbf{algebrae} implementing these functions and a \textbf{grammar} containing parsers that make use of the functions defined in the signature. The concrete program instance combines the algebra with the grammar. The grammar parsers' intermediate results are memoized in a matrix (tabulation parser). A parser usually consist of a tree of:\ul
\item \textbf{Terminal:} operates on a subsequence of input elements and returns either its content or position (or a failure if the sequence does not fit the terminal).
\item \textbf{Filter:} accepts only subsequences matching a certain predicate. The condition is evaluated ahead of its actual content evaluation.
\item \textbf{Or:} expresses alternative between two different parsers and returns their result union.
\item \textbf{Concatenation:} constructs all possible combinations from two subsequences. The subsequences can be of fixed or varying size and concatenation operators might impose restrictions on the subsequences length to be considered.
\item \textbf{Map:} this parser transform its input using a user-defined function. It is typically used to transform a subword into a score that can later be aggregated.
\item \textbf{Aggregation:} the aggregation applies a functions that reduces the list of results, typically minimum or maximum, but the function can be arbitrarily defined.
\item \textbf{Tabulation:} the tabulation's primary function is to store intermediate results and possibly serve as connection point between different parsers.
\ule

@@

The research on ad-hoc implementation has focused on three kind of problems:\ul
\item Genral problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
\item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
\item Biological sequence alignment (Smith-Waterman) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
\ule

>>>>

A \textbf{signature} serves as interface between the grammar, the scoring algebra and the aggregation function, which allows the user to combine one grammar with different algebras or vice versa. Because recurrence relations are expressed by a parsing grammar, ADP makes the candidate structure explicit and hides tabulation indices, thereby preventing potential errors. Finally, since the expression of the dynamic program is formalized and abstracted into a grammar and an algebra, it becomes possible to systematically convert dynamic programming descriptions into efficient recurrences for many-core platforms such as GPUs \cite{adp_gpu}.

@@

Practically, an ADP program is made of 3 components: a \textbf{signature} that define a set of function signatures, one or more \textbf{algebras} implementing these functions and a \textbf{grammar} containing parsers that make use of the functions defined in the signature. The concrete program instance combines the algebra with the grammar. The grammar parsers' intermediate results are memoized in a matrix (tabulation parser). A parser usually consist of a tree of:\ul
\item \textbf{Terminal:} operates on a subsequence of input elements and returns either its content or position (or a failure if the sequence does not fit the terminal).
\item \textbf{Filter:} accepts only subsequences matching a certain predicate. The condition is evaluated ahead of its actual content evaluation.
\item \textbf{Or:} expresses alternative between two different parsers and returns their result union.
\item \textbf{Concatenation:} constructs all possible combinations from two subsequences. The subsequences can be of fixed or varying size and concatenation operators might impose restrictions on the subsequences length to be considered.
\item \textbf{Map:} this parser transform its input using a user-defined function. It is typically used to transform a subword into a score that can later be aggregated.
\item \textbf{Aggregation:} the aggregation applies a functions that reduces the list of results, typically minimum or maximum, but the function can be arbitrarily defined.
\item \textbf{Tabulation:} the tabulation's primary function is to store intermediate results and possibly serve as connection point between different parsers.
\ule

@@

The research on ad-hoc implementation has focused on three kind of problems:\ul
\item General problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
\item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
\item Biological sequence alignment (Smith-Waterman) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
\ule
